Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
more JSS house style bib changes (capitalization)

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, or playing white in
chess~\citep{hankin2020_jss}.  Note that $\lambda < 1$ would correspond to
a decrease in competitive ability.

@@

Equation~\ref{mostgeneral} is very general and may be applied to a
range of situations in which standard Bradley-Terry strengths are
modified in some way.  In the following section, I apply it to
Plackett-Luce likelihoods~\citep{plackett1975} likelihood function for
rank data.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \fct{cheering} which
implements this functionality.

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972}.  The canonical problem
is as follows.  A commuter has a choice of going to work by different
transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972}, for example, presents an approach involving
overlapping ``aspects'' of the elements of choice set;
\cite{mcfadden1980} suggests using a generalized extreme-value model.

>>>>

One generalization of the Bradley-Terry model is due
to~\citet{luce1959_jss}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, or playing white in
chess~\citep{hankin2020_jss}.  Note that $\lambda < 1$ would correspond to
a decrease in competitive ability.

@@

Equation~\ref{mostgeneral} is very general and may be applied to a
range of situations in which standard Bradley-Terry strengths are
modified in some way.  In the following section, I apply it to
Plackett-Luce likelihoods~\citep{plackett1975_jss} likelihood function for
rank data.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975_jss} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \fct{cheering} which
implements this functionality.

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
is as follows.  A commuter has a choice of going to work by different
transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972}, for example, presents an approach involving
overlapping ``aspects'' of the elements of choice set;
\cite{mcfadden1980} suggests using a generalized extreme-value model.
